---
title: "Labor Markets and Unemployment"
author: "Matt Dobra"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
    ioslides_presentation:
        widescreen: TRUE
        incremental: false
        transition: faster
        css: dobraslidestyle.css
      
---
```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(dplyr)
library(kableExtra)
library(ggthemes)
library(ggpubr)
library(DiagrammeR)
library(WDI) # World Development Indicators
library(quantmod) # Gets FRED data
library(scales)
theme_set(theme_classic())
knitr::opts_chunk$set(echo = FALSE, dev.args=list(bg="transparent"))

demand <- tibble(p = c(2,8), q = c(8,2))
demandright <- tibble(p = c(3,9), q = c(9,3))
demandleft <- tibble(p = c(1,7), q = c(7,1))
supply <- tibble(p=c(2,8), q=c(2,8))
supplyright <- tibble(p=c(1,7), q=c(3,9))
supplyleft <- tibble(p = c(3,9), q = c(1, 7))

###Color Themes
color1 <- "#FFFFFF" # white
color2 <- "#00205B" # med blue
color3 <- "#0A863D" # green
color4 <- "#99999A" # gray
color5 <- "#041C2C" # Navy

######
# 
# demand <- tibble(p= c(2,8), q=c(8,2))
# demandright <- tibble(p = c(3,9), q = c(9,3))
# demandleft <- tibble(p = c(1,7), q = c(7,1))
# supply <- tibble(p=c(2,8), q=c(2,8))
# supplyright <- tibble(p=c(1,7), q=c(3,9))
# supplyleft <- tibble(p = c(3,9), q = c(1, 7))


## Create Recession dataframe
getSymbols("USREC",src="FRED")
USREC <- USREC["1855/"]
start <- index(USREC[which(diff(USREC$USREC)==1)])
end   <- index(USREC[which(diff(USREC$USREC)==-1)-1])
if(length(end)<length(start)){
end <- c(end, Sys.Date())    
}
recession.df <- data.frame(start=start, end=end)
rm(start)
rm(end)
rm(USREC)

```

# Measuring Unemployment
## Unemployment

There are a variety of types of unemployment, and all types are not created equal.

Some unemployment is an unavoidable and necessary consequence of economic growth, as economic progress inevitably renders some jobs obsolete.

- Employment in fields like farming have declined significantly over the past 150 years, and some jobs have essentially disappeared (telephone switchboard operators, wagonmakers, etc.).
- These jobs are destroyed because of the appearance of new jobs in areas like IT server technician and automotive engineer.

## Unemployment

<div style="float:right; width:70%;">
```{r get_unrate, include = FALSE}
getSymbols("UNRATE", src = "FRED", return.class = "data.frame") # U3 Unemployment Rate
unrate <- UNRATE
rm(UNRATE)
unrate <- rename(unrate, unrate = UNRATE)
unrate <- rownames_to_column(unrate, "date")
unrate$date <- as.Date(unrate$date)

curunemp <- unrate[length(unrate$unrate), ]

lubridate::month(as.Date(curunemp$date[1]), label = TRUE, abbr = FALSE)

```

```{r warning = FALSE, out.width = "100%", out.extra='style="float:right; padding:1px"'}

ggplot() +
    geom_line(data = unrate, aes(x = date, y = unrate), color = color2, size = 1.2) + 
    geom_rect(data = recession.df, aes(xmin = start, xmax = end, ymin = -Inf, ymax = +Inf), fill = color5, alpha = .3) +
     theme_classic()+
    scale_y_continuous(breaks = seq(0,20, by = 2), 
                       limits = c(0,NA),
                       expand = c(0,0)) +
    scale_x_date(limits = as.Date(c("1948-01-01", Sys.Date())),
        expand = c(0,0),   
        labels = date_format("%Y"),
        breaks = function(x)
            seq.Date(
                from = as.Date("1940-01-01"),
                to = as.Date(Sys.Date()),
                by = "10 years"
            )) +
    labs(
        title = "Unemployment",
        subtitle = "U-3 unemployment, monthy data since 1948",
        x = "Date",
        y = "Unemloyment Rate",
        caption = "Data from FRED, created by Matt Dobra @mattdobra"
    ) 
```
</div>

<div style="float:left; width:30%;">

<br> 
The key measure that economists look at in the US is the U-3 Unemployment rate.

The unemployment rate as of `r lubridate::month(as.Date(curunemp$date[1]), label = TRUE, abbr = FALSE)`, `r lubridate::year(as.Date(curunemp$date[1]))` is `r curunemp$unrate[1]`%.

</div>

## Unemployment

To calculate the unemployment rate, the population must first be subdivided into a number of smaller groups.


## Unemployment

<div style="float:right; width:70%;">
```{r get_civpop, include = FALSE}
getSymbols("CNP16OV", src = "FRED", return.class = "data.frame") # U3 Unemployment Rate
civpop <- CNP16OV
rm(CNP16OV)
civpop <- rename(civpop, civpop = CNP16OV)
civpop <- rownames_to_column(civpop, "date")
civpop$date <- as.Date(civpop$date)

curcivpop <- civpop[length(civpop$civpop), ]
```

```{r warning = FALSE, out.width = "100%", out.extra='style="float:right; padding:1px"'}

ggplot() +
    geom_line(data = civpop, aes(x = date, y = civpop), color = color2, size = 1.2) + 
    geom_rect(data = recession.df, aes(xmin = start, xmax = end, ymin = -Inf, ymax = +Inf), fill = color5, alpha = .3) +
     theme_classic()+
    scale_y_continuous(breaks = seq(0,300000, by = 25000), 
                       limits = c(0,NA),
                       expand = c(0,10000)) +
    scale_x_date(limits = as.Date(c("1948-01-01", Sys.Date())),
        expand = c(0,0),   
        labels = date_format("%Y"),
        breaks = function(x)
            seq.Date(
                from = as.Date("1940-01-01"),
                to = as.Date(Sys.Date()),
                by = "10 years"
            )) +
    labs(
        title = "Population",
        subtitle = "US civilian population, monthy data since 1948",
        x = "Date",
        y = "Millions of persons",
        caption = "Data from FRED, created by Matt Dobra @mattdobra"
    ) 
```
</div>

<div style="float:left; width:30%;">

**Civilian Noninstitutional Population**: Number of people in the US or DC, over 16, not inmates or institutionalized, or in the Armed Forces.

As of `r lubridate::month(as.Date(curcivpop$date[1]), label = TRUE, abbr = FALSE)`, `r lubridate::year(as.Date(curcivpop$date[1]))`, this figure for the US is `r round(curcivpop$civpop[1]/1000, 2)` million.

</div>

## Unemployment

<div style="float:right; width:70%;">
```{r get_labforce, include = FALSE}
getSymbols("CLF16OV", src = "FRED", return.class = "data.frame") # Labor Force
labforce <- CLF16OV
rm(CLF16OV)
labforce <- rename(labforce, labforce = CLF16OV)
labforce <- rownames_to_column(labforce, "date")
labforce$date <- as.Date(labforce$date)

curlabforce <- labforce[length(labforce$labforce), ]
```

```{r warning = FALSE, out.width = "100%", out.extra='style="float:right; padding:1px"'}

ggplot() +
    geom_line(data = labforce, aes(x = date, y = labforce), color = color2, size = 1.2) + 
    geom_rect(data = recession.df, aes(xmin = start, xmax = end, ymin = -Inf, ymax = +Inf), fill = color5, alpha = .3) +
     theme_classic()+
    scale_y_continuous(breaks = seq(0,300000, by = 20000), 
                       limits = c(0,NA),
                       expand = c(0,10000)) +
    scale_x_date(limits = as.Date(c("1948-01-01", Sys.Date())),
        expand = c(0,0),   
        labels = date_format("%Y"),
        breaks = function(x)
            seq.Date(
                from = as.Date("1940-01-01"),
                to = as.Date(Sys.Date()),
                by = "10 years"
            )) +
    labs(
        title = "Labor Force",
        subtitle = "Total labor force, data since 1948",
        x = "Date",
        y = "Persons in Labor Force",
        caption = "Data from FRED, created by Matt Dobra @mattdobra"
    ) 
```
</div>

<div style="float:left; width:30%;">

**Labor Force**: Civilian noninstitutionalized population, less retirees, students, homemakers, and people just not looking for work.  

As of `r lubridate::month(as.Date(curlabforce$date[1]), label = TRUE, abbr = FALSE)`, `r lubridate::year(as.Date(curlabforce$date[1]))`, the US labor force numbered  `r round(curlabforce$labforce[1]/1000, 2)` million people.

</div>

## Unemployment

<div style="float:right; width:70%;">
```{r get_partrate, include = FALSE}
getSymbols("CIVPART", src = "FRED", return.class = "data.frame") # Labor Force
partrate <- CIVPART
rm(CIVPART)
partrate <- rename(partrate, partrate = CIVPART)
partrate <- rownames_to_column(partrate, "date")
partrate$date <- as.Date(partrate$date)

curpartrate <- partrate[length(partrate$partrate), ]
```

```{r warning = FALSE, out.width = "100%", out.extra='style="float:right; padding:1px"'}

ggplot() +
    geom_line(data = partrate, aes(x = date, y = partrate), color = color2, size = 1.2) + 
    geom_rect(data = recession.df, aes(xmin = start, xmax = end, ymin = -Inf, ymax = +Inf), fill = color5, alpha = .3) +
     theme_classic()+
    scale_y_continuous(breaks = seq(0,100, by = 1), 
                       limits = c(NA,NA),
                       expand = c(NA,1)) +
    scale_x_date(limits = as.Date(c("1948-01-01", Sys.Date())),
        expand = c(0,0),   
        labels = date_format("%Y"),
        breaks = function(x)
            seq.Date(
                from = as.Date("1940-01-01"),
                to = as.Date(Sys.Date()),
                by = "10 years"
            )) +
    labs(
        title = "Labor Force Participation",
        subtitle = "US labor force participation rate, data since 1948",
        x = "Date",
        y = "Percent in Labor Force",
        caption = "Data from FRED, created by Matt Dobra @mattdobra"
    ) 
```
</div>

<div style="float:left; width:30%;">

**Labor Force Participation Rate**: The percentage of the civilian population that is in the labor force.  

As of `r lubridate::month(as.Date(curpartrate$date[1]), label = TRUE, abbr = FALSE)`, `r lubridate::year(as.Date(curpartrate$date[1]))`, the US labor force participation rate was  `r curpartrate$partrate[1]`\%.

</div>

## Unemployment

```{r get_partratebyage, include = FALSE}
getSymbols("LNS11300012", src = "FRED", return.class = "data.frame") # Participation Rage 16 - 19
getSymbols("LNS11300060", src = "FRED", return.class = "data.frame") # Participation Rage 25 - 54
getSymbols("LNS11324230", src = "FRED", return.class = "data.frame") # Participation Rage 55 +

LNS11300012 <- rownames_to_column(LNS11300012, "date")
LNS11300012$date <- as.Date(LNS11300012$date)

LNS11300060 <- rownames_to_column(LNS11300060, "date")
LNS11300060$date <- as.Date(LNS11300060$date)

LNS11324230 <- rownames_to_column(LNS11324230, "date")
LNS11324230$date <- as.Date(LNS11324230$date)


LNS11300012 <- rename(LNS11300012, partrate = LNS11300012) %>% mutate(group = "16-19") 
LNS11300060 <- rename(LNS11300060, partrate = LNS11300060) %>% mutate(group = "25-54")
LNS11324230 <- rename(LNS11324230, partrate = LNS11324230) %>% mutate(group = "55+")

partratebyage <- LNS11300012 %>% 
    bind_rows(LNS11300060) %>% 
    bind_rows(LNS11324230)


```

<div style="float:left; width:30%;">

The labor force participation rate is heavily influenced by demographics.

it is also influenced by incentives -- high tax rates and generous welfare benefits reduce the participation rate. 

</div>

<div style="float:right; width:70%;">


```{r warning = FALSE, out.width = "100%", out.extra='style="float:right; padding:1px"'}

ggplot() +
    geom_line(data = partratebyage, aes(x = date, y = partrate, color = group), size = 1.2) + 
    geom_rect(data = recession.df, aes(xmin = start, xmax = end, ymin = -Inf, ymax = +Inf), fill = color5, alpha = .3) +
     theme_classic()+
    scale_y_continuous(breaks = seq(0,100, by = 5), 
                       limits = c(NA,NA),
                       expand = c(NA,1)) +
    scale_x_date(limits = as.Date(c("1948-01-01", Sys.Date())),
        expand = c(0,0),   
        labels = date_format("%Y"),
        breaks = function(x)
            seq.Date(
                from = as.Date("1940-01-01"),
                to = as.Date(Sys.Date()),
                by = "10 years"
            )) +
    labs(
        title = "Labor Force Participation",
        subtitle = "US labor force participation rate by age group, data since 1948",
        x = "Date",
        y = "Percent in Labor Force",
        caption = "Data from FRED, created by Matt Dobra @mattdobra",
        color = "Age Group"
    ) +
    scale_color_manual(values = c(color2, color3, color4))
    
```
</div>



## Unemployment

A person is **unemployed** only if he or she is in the labor force, is actively looking for work, but cannot find a job.

The **unemployment rate** is then given by:

$Unemployment \: Rate = \frac{\# \: of \: Unemployed}{Labor \: Force}$

```{r get_unemploy, include = FALSE}
getSymbols("UNEMPLOY", src = "FRED", return.class = "data.frame") # Labor Force
unemploy <- UNEMPLOY
rm(UNEMPLOY)
unemploy <- rename(unemploy, unemploy = UNEMPLOY)
unemploy <- rownames_to_column(unemploy, "date")
unemploy$date <- as.Date(unemploy$date)

curunemploy <- unemploy[length(unemploy$unemploy), ]
```

As of `r lubridate::month(as.Date(curlabforce$date[1]), label = TRUE, abbr = FALSE)`, `r lubridate::year(as.Date(curlabforce$date[1]))`, the US labor force numbered `r round(curlabforce$labforce[1]/1000, 2)` million people, of which `r round(curunemploy$unemploy[1]/1000, 2)` million were unemployed.  This implies an unemployment rate of:

$`r round(100*curunemploy$unemploy[1]/curlabforce$labforce[1],1)`\% = \frac{`r round(curunemploy$unemploy[1]/1000, 2)` \: million }{`r round(curlabforce$labforce[1]/1000,2)` \: million}$

## Unemployment

From an individual perspective, unemployment, especially if it is long term unemployment, can be both financial and psychologically devastating.

From a economy-wide perspective, unemployment means that the economy is under-performing.

While the unemployment rate is arguably the *best* indicator of the labor market, it is not the whole story.

- The unemployment rate misses discouraged and underemployed workers.

## Unemployment

<div style = "float: left; width:30%">

A **discouraged worker** is someone who is not currently looking for a job because they believe they will not find one, but has looked for work within the past year.  

Discouraged workers are included in the U-4, which is typically a little higher than the U-3.

</div>

<div style = "float: right; width:70%;">

```{r get_u4, include = FALSE}
getSymbols("U4RATE", src = "FRED", return.class = "data.frame") # U3 Unemployment Rate
unrate4 <- U4RATE
rm(U4RATE)
unrate4 <- rename(unrate4, unrate4 = U4RATE)
unrate4 <- rownames_to_column(unrate4, "date")
unrate4$date <- as.Date(unrate4$date)

curunemp4 <- unrate4[length(unrate4$unrate4), ]

lubridate::month(as.Date(curunemp4$date[1]), label = TRUE, abbr = FALSE)

```

```{r warning = FALSE, out.width = "100%", out.extra='style="float:right; padding:1px"'}

ggplot() +
    geom_line(data = unrate, aes(x = date, y = unrate), color = color2, size = 1.2) + 
    geom_line(data = unrate4, aes(x = date, y = unrate4), color = color3, size = 1.2) + 
    annotate(geom = "text", x = as.Date("2011-01-01"), y = 11.5, label = "U-4 Rate", color = color3, size = 5)+
    annotate(geom = "text", x = as.Date("2011-01-01"), y = 7.5, label = "U-3 Rate", color = color2, size = 5)+
    geom_rect(data = recession.df, aes(xmin = start, xmax = end, ymin = -Inf, ymax = +Inf), fill = color5, alpha = .3) +
     theme_classic()+
    scale_y_continuous(breaks = seq(0,20, by = 2), 
                       limits = c(0,NA),
                       expand = c(0,0)) +
    scale_x_date(limits = as.Date(c("1994-01-01", Sys.Date())),
        expand = c(0,0),   
        labels = date_format("%Y"),
        breaks = function(x)
            seq.Date(
                from = as.Date("1990-01-01"),
                to = as.Date(Sys.Date()),
                by = "5 years"
            )) +
    labs(
        title = "Unemployment",
        subtitle = "U-4 unemployment, monthy data since 1994",
        x = "Date",
        y = "Unemloyment Rate",
        caption = "Data from FRED, created by Matt Dobra @mattdobra"
    ) 
```

## Unemployment

The broadest measure of unemployment is the U-6, which includes discouraged, marginally attached, and underemployed workers.

A **marginally attached worker** someone who has either looked for work or held a job in the past year, but is not currently looking.  

U-3 considers somebody employed if they worked as little as one hour.  However, many people working part time would prefer to work full time but cannot find a full time job.  These people are classified as **underemployed workers**.  


## Unemployment

```{r get_u6, include = FALSE}
getSymbols("U6RATE", src = "FRED", return.class = "data.frame") # U3 Unemployment Rate
unrate6 <- U6RATE
rm(U6RATE)
unrate6 <- rename(unrate6, unrate6 = U6RATE)
unrate6 <- rownames_to_column(unrate6, "date")
unrate6$date <- as.Date(unrate6$date)

curunemp6 <- unrate6[length(unrate6$unrate6), ]

lubridate::month(as.Date(curunemp6$date[1]), label = TRUE, abbr = FALSE)

```
<div style = "float: left; width:30%">

While the U-3 unemployment rate in `r lubridate::month(as.Date(curunemp$date[1]), label = TRUE, abbr = FALSE)`, `r lubridate::year(as.Date(curunemp$date[1]))` was `r curunemp$unrate[1]`%, the U-6 unemployment rate was `r curunemp6$unrate6[1]`%.

The gap between the U-6 and the U-3 is large and grew following the great recession.

</div>

<div style = "float: right; width:70%;">



```{r warning = FALSE, out.width = "100%", out.extra='style="float:right; padding:1px"'}

ggplot() +
    geom_line(data = unrate, aes(x = date, y = unrate), color = color2, size = 1.2) + 
    geom_line(data = unrate6, aes(x = date, y = unrate6), color = color3, size = 1.2) + 
    annotate(geom = "text", x = as.Date("2004-01-01"), y = 11.5, label = "U-6 Rate", color = color3, size = 5)+
    annotate(geom = "text", x = as.Date("2004-01-01"), y = 7.5, label = "U-3 Rate", color = color2, size = 5)+
    geom_rect(data = recession.df, aes(xmin = start, xmax = end, ymin = -Inf, ymax = +Inf), fill = color5, alpha = .3) +
     theme_classic()+
    scale_y_continuous(breaks = seq(0,30, by = 2), 
                       limits = c(0,NA),
                       expand = c(0,0)) +
    scale_x_date(limits = as.Date(c("1994-01-01", Sys.Date())),
        expand = c(0,0),   
        labels = date_format("%Y"),
        breaks = function(x)
            seq.Date(
                from = as.Date("1990-01-01"),
                to = as.Date(Sys.Date()),
                by = "5 years"
            )) +
    labs(
        title = "Unemployment",
        subtitle = "U-6 unemployment, monthy data since 1994",
        x = "Date",
        y = "Unemloyment Rate",
        caption = "Data from FRED, created by Matt Dobra @mattdobra"
    ) 
```



# Types of Unemployment
## Types of Unemployment

The three main types of unemployment are:

- Frictional Unemployment
- Structural Unemployment
- Cyclical Unemployment

## Frictional Unemployment

**Frictional Unemployment** is (typically) short term unemployment caused by the difficulties that exist in matching potential employees with jobs.

Many people who are frictionally unemployed are unemployed by choice, as this includes people leaving their jobs to find new ones.  It also includes recent college graduates, stay-at-home parents returning to the workforce, etc.

For the most part, frictional unemployment is good for the economy.


## Frictional Unemployment

Economist Joseph Schumpeter coined the term **creative destruction** to describe the constant process of product and process innovation that replaces old products and processes. 

Schumpeter called creative destruction "the essential fact of capitalism"

Innovation and competition drive economic growth, creating new jobs and destroying old jobs.

Official government figures mask the creative destruction process.  

- For example, the number people focus on in the jobs report is the number of new jobs in a month.  In August 2018, this number was 78,000.  

- The 78,000 figure is arrived at by taking the difference between the number of new hires, 5.784 million, and the number of job separations, 5.706 million.  

## Structural Unemployment

**Structural unemployment** is persistent, typically long-term unemployment that is caused by long-lasting shocks or permanent features of an economy that prevent the labor market from working effectively.

Structural unemployment can result from a number of economic forces:

- economy-wide shocks, like oil shocks (or pandemics?).
- globalization
- new IT
- shifting from manufacturing to services

## Structural Unemployment

Structural unemployment has significant economic and social costs.

- Lowered economic output
- Psychological costs on the unemployed, such as elevated suicide rates and reduced happiness.

## Structural Unemployment

<div style = "float:left; width:40%;">

Structural unemployment is a larger problem in countries with strong labor market regulations. 

A 2013 OECD study measured the strictness of labor market regulations.  

Having more and stronger regulations is positively correlated with duration of unemployment
</div>

<div style = "float:right; width:60%;">


```{r include = FALSE}
trgtyear <- 2013
prot <- OECD::get_dataset("EPL_OV", start_time = trgtyear, end_time = trgtyear)
prot1 <- prot %>% 
    filter(SERIES == "EPRC_V3") %>% 
    select(COUNTRY, ObsValue) %>% 
    rename(prot = ObsValue)
filter_list <- list("", "MW", "900000", "UN5")
dur <- OECD::get_dataset("DUR_I", start_time = trgtyear, end_time = trgtyear, filter = filter_list)
dur1 <- dur %>% 
    select(COUNTRY, ObsValue) %>% 
    rename(duration = ObsValue)

dat <- prot1 %>% 
    inner_join(dur1)

```

```{r warning = FALSE, message = FALSE}
dat$prot <- as.numeric(dat$prot)
dat$duration <- as.numeric(dat$duration)
ggplot(data = dat, aes(x = prot, y = duration, label = COUNTRY)) +
    ggrepel::geom_text_repel() +
    geom_smooth(method = lm, color = color3, fill = color4) +
    # scale_y_continuous(limits = c(0,70)) +
    # scale_x_continuous(limits = c(0,4)) +
    labs(
        title = "Labor Market Protection",
        subtitle = "Stricter labor market regulation results in longer average unemployment duration",
        x = "Index of Labor Market Regulations",
        y = "Average Length of Unemployment (weeks)",
        caption = "Data from OECD, created by Matt Dobra @mattdobra"
    ) 

```
</div>

## Structural Unemployment

Labor market regulations generally have the following effects:

- Create insurance for workers, particularly those workers who have full-time jobs.
- Make it harder to find a full-time job.
- Reduce the efficiency of labor markets.
- Increase the duration of unemployment.
- Increase the amount of unemployment, particularly among young, minority, or "risky" workers. 
    - e.g. ex-convicts, people you know little about, high variance workers...

## Structural Unemployment


<div style="float: left; width: 50%;">

Minimum wages also increase unemployment.  

The higher the minimum wage is relative to the market wage, the greater the unemployment created.

This implies that countries with higher minimum wages (relative to market wages) should have more unemployment.

</div>

```{r fig.width=5,fig.height=5, out.extra='style="float:right; padding:1px"'}
demand2 <- tibble(p= c(1,10), q= c(9,0))
supply2 <- tibble(p= c(0,9), q= c(0,9))

ggplot(mapping = aes(x = q, y = p)) + 
  ## Base Graph Begin**
  geom_path(data = demand2, color = color2, size = 1.2) + 
  annotate("text", x=9.2, y=.9, label = "D", color = color2, size = 6, parse = TRUE) +
  geom_path(data = supply2, color = color3, size = 1.2) + 
  annotate("text", x=9.2, y=9.1, label = "S", color = color3, size = 6, parse = TRUE) +
  geom_bracket(xmin = 3, xmax = 7, y.position = 7.5, label = "Unemployment", tip.length = .05) +
  coord_equal() +
  theme_classic(base_family = "sans") +
  theme(plot.title = element_text(size = rel(1.4))) +
  ## End Base Graph ##

    ## Graph Labels -- Change title and subtitle as needed ##
  labs(x = "Quantity", y = "Wage",
       title = "Minimum Wage",
       subtitle = expression(paste("The distance between ", Q[S], "&", Q[D], " represents unemployment"))) + # Subtitle
  
  ##editing the x and y axes for the equilibria.  If dots aren't needed, still need to keep limit and expand##
  scale_x_continuous(expand = c(0,0), limit = c(0,10), breaks = c(5,3,7), labels = expression(Q[1], Q[D], Q[S])) + ## Set breaks = equilibrium Q
  scale_y_continuous(expand = c(0,0), limit = c(0,10), breaks = c(5,7), labels = expression(P[1], P[mw])) + ## Set breaks = equilibrium P
  
  ## P/Q Arrows (delete if no changes, edit points otherwise. 
  ## Leave .4 from axes and 0.2 from start/end point 
  annotate("segment", x = .4, xend = .4, y = 5.2, yend = 6.8,
           arrow = arrow(length = unit(.5, "lines")), colour = color4, size = 1) +    
  
  
  ## Dashed Lines for initial equilibrium
  annotate("segment", x=0, xend = 5, y = 5, yend = 5, linetype = "dotted") + # horizontal line
  annotate("segment", x=5, xend = 5, y = 0, yend = 5, linetype = "dotted") + # vertical line
  annotate("segment", x=0, xend = 7, y = 7, yend = 7, linetype = "dotted") + # horizontal line
  annotate("segment", x=3, xend = 3, y = 0, yend = 7, linetype = "dotted") + # vertical line
  annotate("segment", x=7, xend = 7, y = 0, yend = 7, linetype = "dotted") # vertical line
```

## Structural Unemployment

The **Kaitz index** is a measure of how high the minimum wage is in any given area.

$Kaitz\;Index=\frac{Minimum\:Wage}{Median\:Wage}$

The median wage is the wage such that half of all workers earn more the median and half of all workers earn less than the median.

## Structural Unemployment

The Kaitz index is often use as a shorthand to estimate the potential harm of a minimum wage.  

- If Kaitz<40%, the general consensus is that the minimum wage is too low to have a noticeable unemployment effect.
- If Kaitz>60%, most economists believe that the minimum wage will have a large unemployment effect.
- There is no general consensus on the 40%-60% range.

## Structural Unemployment

<div style = "float:left; width:40%;">

The Kaitz index varies widely across OECD countries.

Countries tend to lie in the range between 40% and 60%, but not all do.

Note that this graph is based on federal minimum wages, and local minimum wages may be higher.

</div>

<div style = "float:right; width:60%;">


```{r include = FALSE}
trgtyear <- lubridate::year(as.Date(Sys.time())) - 2
filter_list <- list("", "MEDIAN")
min2ave <- OECD::get_dataset("MIN2AVE", start_time = trgtyear, end_time = trgtyear, filter = filter_list)

min2ave2 <- min2ave %>% 
  rename(kaitz = ObsValue) %>% 
  select(COUNTRY, kaitz) %>% 
  mutate(country = countrycode::countrycode(COUNTRY, origin = "wb", destination = "country.name")) %>% 
  arrange(kaitz)
min2ave2$kaitz <- as.numeric(min2ave2$kaitz)

```

```{r warning = FALSE, message = FALSE, fig.height = 6}
min2ave2 %>% 
    mutate(kaitz = kaitz / 100) %>% 
    ggplot() +
  geom_col(aes(x = reorder(country, kaitz), y = kaitz), fill = color2) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent, breaks = c(0, .20, .40, .60, .80, 1.00), expand = c(0,NA)) +
 # theme(axis.text.y = element_text(size = 11)) +
  labs(title = "Kaitz Index",
       subtitle = paste("Ratio of minimum wage to median wage in ", trgtyear),
       x = "Kaitz Index",
       y = "",
       caption = "Data from OECD, created by Matt Dobra @mattdobra")
  


```

## Structural Unemployment

<div style = "float:left; width:40%;">

<br>
<br>
There is generally an increasing relationship between the Kaitz Index and unemployment rates.


</div>

<div style = "float:right; width:60%;">


```{r include = FALSE}
trgtyear <- lubridate::year(as.Date(Sys.time())) - 2
filter_list <- list("", "MW", "900000", "UR")
unemp <- OECD::get_dataset("LFS_SEXAGE_I_R", start_time = trgtyear, end_time = trgtyear, filter = filter_list)

unemp2 <- unemp %>% 
  rename(unemp = ObsValue) %>% 
  select(COUNTRY, unemp) %>% 
  inner_join(min2ave2)

stargazer::stargazer(lm(unemp ~ kaitz, data = unemp2), type = "text")
unemp2$unemp <- as.numeric(unemp2$unemp)
```

```{r warning = FALSE, message = FALSE}
ggplot(data = unemp2, aes(x = kaitz, y = unemp, label = COUNTRY)) +
    ggrepel::geom_text_repel() +
    labs(
        title = "Kaitz Index and Unemployment",
        subtitle = "",
        x = "Kaitz Index",
        y = "Unemployment Rate (%)",
        caption = "Data from OECD, created by Matt Dobra @mattdobra"
    ) +
  scale_x_continuous(labels =scales::percent_format(accuracy = 1))

```

## Structural Unemployment

There are also policies that can reduce structural unemployment:

**At-Will Employment** refers to a labor market in which a worker can be fired or quit at any time for any reason, so long as that reason is not illegal (discrimination, etc.).  

**Active Labor Market Policies** are (typically) government programs designed to help unemployed workers find jobs by:

- enhancing their motivation and incentives to find work
- improving their job readiness and skills through education or job retraining programs
- improving their job search skills and assisting with their job search

## Cyclical Unemployment

**Cyclical Unemployment** is unemployment that is correlated with the business cycle.

Generally speaking, unemployment increases during a recession.

When GDP falls, firms fire workers, thus increasing unemployment.

## Cyclical Unemployment

<div style="float:right; width:70%;">

```{r warning = FALSE, out.width = "100%", out.extra='style="float:right; padding:1px"'}

ggplot() +
    geom_line(data = unrate, aes(x = date, y = unrate), color = color2, size = 1.2) + 
    geom_rect(data = recession.df, aes(xmin = start, xmax = end, ymin = -Inf, ymax = +Inf), fill = color5, alpha = .3) +
     theme_classic()+
    scale_y_continuous(breaks = seq(0,20, by = 2), 
                       limits = c(0,NA),
                       expand = c(0,0)) +
    scale_x_date(limits = as.Date(c("1948-01-01", Sys.Date())),
        expand = c(0,0),   
        labels = date_format("%Y"),
        breaks = function(x)
            seq.Date(
                from = as.Date("1940-01-01"),
                to = as.Date(Sys.Date()),
                by = "10 years"
            )) +
    labs(
        title = "Unemployment",
        subtitle = "U-3 unemployment, monthy data since 1948",
        x = "Date",
        y = "Unemloyment Rate",
        caption = "Data from FRED, created by Matt Dobra @mattdobra"
    ) 
```
</div>

<div style="float:left; width:30%;">

Consider again the U-3 unemployment graph.  

Unemployment rates generally spike during recessions and slowly fall afterwards.

</div>

## Cyclical Unemployment

<div style = "float:left; width:50%;">

<br>

Cyclical unemployment can change dramatically in a very short time period.  

For example, consider the unemployment rate for the months immediately surrounding the Covid-19 crisis in 2020:

</div>

<div style = "float:right; width:50%;">

<br>

```{r}

unrate %>% 
    filter(date >= as.Date("2020-01-01")) %>% 
    filter(date <= as.Date("2020-06-01")) %>% 
    mutate(date = lubridate::month(date, label = TRUE, abbr = FALSE)) %>% 
    rename(Month = date) %>% 
    rename(Unemployment = unrate) %>% 
    knitr::kable(format = "html", table.attr = "class=\"rmdtable\"", col.names = c("Month",
                           "Unemploment Rate")) %>% 
    column_spec(1, background = color2, color = color1)

```

</div>

## Cyclical Unemployment

<div style="float:right; width:70%;">

```{r warning = FALSE, out.width = "100%", out.extra='style="float:right; padding:1px"'}

ggplot() +
    geom_line(data = unrate, aes(x = date, y = unrate), color = color2, size = 1.2) + 
    geom_rect(data = recession.df, aes(xmin = start, xmax = end, ymin = -Inf, ymax = +Inf), fill = color5, alpha = .3) +
     theme_classic()+
    scale_y_continuous(breaks = seq(0,20, by = 2), 
                       limits = c(0,NA),
                       expand = c(0,0)) +
    scale_x_date(limits = as.Date(c("2020-01-01", "2021-12-01")),
        expand = c(0,0),   
        labels = date_format("%b %Y"),
        breaks = function(x)
            seq.Date(
                from = as.Date("1940-01-01"),
                to = as.Date(Sys.Date()),
                by = "3 months"
            )) +
    labs(
        title = "Covid and Cyclical Unemployment",
        subtitle = "U-3 unemployment, 2020 and 2021",
        x = "Month and Year",
        y = "Unemloyment Rate",
        caption = "Data from FRED, created by Matt Dobra @mattdobra"
    ) 
```
</div>

<div style="float:left; width:30%;">

Zooming in on the U-3 graph to 2020 and 2021 shows the cyclical unemployment impact of Covid.  

Even with the Covid recovery being remarkably quick, the recovery is still much slower than the initial shock.

</div>



## Natural Rate of Unemployment

The **Natural Rate of Unemployment** is calculated by summing the frictional and structural rates of unemployment.

The natural rate changes very slowly
- The factors that influence structural and frictional unemployment change slowly


This is unlike the regular unemployment rate which includes cyclical unemployment

## Natural Rate of Unemployment


<div style="float:right; width:70%;">

```{r get_nrou, include = FALSE}
getSymbols("NROU", src = "FRED", return.class = "data.frame") # U3 Unemployment Rate
natrate <- NROU
rm(NROU)
natrate <- rename(natrate, natrate = NROU)
natrate <- rownames_to_column(natrate, "date")
natrate$date <- as.Date(natrate$date)

natrate <- natrate %>% 
    filter(date<as.Date(Sys.Date()))

```

```{r warning = FALSE, out.width = "100%", out.extra='style="float:right; padding:1px"'}

ggplot() +
    geom_line(data = unrate, aes(x = date, y = unrate), color = color2, size = 1.2) + 
    annotate(geom = "text", x = as.Date("1968-01-01"), y = 3, label = "Unemployment rate", color = color2) +
    geom_line(data = natrate, aes(x = date, y = natrate), color = color3, size = 1.2) +
    annotate(geom = "text", x = as.Date("1968-01-01"), y = 6.7, label = "Natural rate", color = color3) +
    geom_rect(data = recession.df, aes(xmin = start, xmax = end, ymin = -Inf, ymax = +Inf), fill = color5, alpha = .3) +
     theme_classic()+
    scale_y_continuous(breaks = seq(0,20, by = 2), 
                       limits = c(0,NA),
                       expand = c(0,0)) +
    scale_x_date(limits = as.Date(c("1949-01-01", Sys.Date())),
        expand = c(0,0),   
        labels = date_format("%Y"),
        breaks = function(x)
            seq.Date(
                from = as.Date("1940-01-01"),
                to = as.Date(Sys.Date()),
                by = "10 years"
            )) +
    labs(
        title = "Natural Rate of Unemployment",
        subtitle = "Natural rate and U-3 unemployment, monthy data since 1949",
        x = "Date",
        y = "Unemloyment Rate",
        caption = "Data from FRED, created by Matt Dobra @mattdobra"
    ) 
```
</div>

<div style="float:left; width:30%;">

<br> 

The natural rate is far more stable than the U-3 rate.

Some economists believe that the risk of inflation increases when unemployment falls below the natural rate.

</div>

